---
title: "Raw Counts Analysis"
author: "Aubrey Odom-Mabey"
date: "1/19/2022"
output:
  html_document:
    toc: true
    theme: united
    toc_depth: 6
    number_sections: true
    code_folding: hide
---
## Setup
```{r setup}
suppressPackageStartupMessages({
  # General use
  library(tidyverse)
  library(knitr)
  library(magrittr)
  # Data formats
  library(animalcules)
  library(MultiAssayExperiment)
  library(SummarizedExperiment)
  # Modeling
  library(lme4)
  library(geepack)
})
```

## Initial Data Extraction
```{r plotting}
# Read in MAE of data
OG_dat <- readRDS("data/animalculesFinalHIV.rds")
dat <- OG_dat[["MicrobeGenetics"]]

# Extract metadata for infants only
microbe <- dat[, which(dat$MothChild == "Infant")]
sam_table <- as.data.frame(colData(microbe))

# Both mothers and infants
microbe_mi <- dat
tax_table <- as.data.frame(SummarizedExperiment::rowData(microbe))

# Extract metadata, taxonomic info, and counts
tax_table_mi <- as.data.frame(SummarizedExperiment::rowData(microbe_mi))
sam_table_mi <- as.data.frame(SummarizedExperiment::colData(microbe_mi))
counts_table_mi <- as.data.frame(SummarizedExperiment::assay(
  microbe_mi, "MGX"))[, rownames(sam_table_mi)]
```

## Statistics
The following statistics are calculated specifically for the infants in this study (HIV-E vs control infants).

### Information for mothers' and infants' samples (using raw reads)

```{R sample info on ages}
sam_table_mi %>%
  filter(MothChild == "Infant") %>%
  dplyr::group_by(timepoint) %>%
  dplyr::summarise("Average Infant Age" = round(mean(Age)),
                   "SD of Infant Age" = round(sd(Age)))
```

```{r table1}
sam_table_mi %>%
  dplyr::rename("HIV Status" = HIVStatus) %>%
  mutate("Timepoint" = factor(timepoint)) %>%
  table1::table1(~ Timepoint + `HIV Status` | factor(MothChild),
               data = .)
```

### Read counts
```{R}
reads_tbl <- counts_table_mi %>%
  as_tibble() %>%
  mutate(species = tax_table_mi$species) %>%
  relocate(species) %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "Sample", values_to = "Abundance") %>%
  left_join(., sam_table_mi, by = "Sample") %>%
  group_by(HIVStatus, MothChild, Sample) %>%
  dplyr::summarise("Total Reads" = sum(`Abundance`)) %>%
  arrange(desc(`Total Reads`))

reads_tbl %>%
  ungroup(HIVStatus) %>%
    dplyr::summarise(med_reads = median(`Total Reads`),
                     mean_reads = mean(`Total Reads`),
                     sd_reads = sd(`Total Reads`),
                     min_reads = min(`Total Reads`),
                     max_reads = max(`Total Reads`),
                     num_total = n())

reads_tbl %>%
    dplyr::summarise(med_reads = median(`Total Reads`),
                     mean_reads = mean(`Total Reads`),
                     sd_reads = sd(`Total Reads`),
                     min_reads = min(`Total Reads`),
                     max_reads = max(`Total Reads`),
                     num_total = n())
```

#### Comparing read counts
```{R}
reads_tbl %>%
  as.data.frame() %>%
  t.test(`Total Reads` ~ MothChild, data = ., paired = FALSE,
       alternative = "greater")
```

### Infant ages at timepoints
```{R}
sam_table %>%
  group_by(timepoint) %>%
  dplyr::summarise(median(`Age`))
# Number of mother samples
# Number of samples' reads
```

## Percent abundances of taxa
```{R create long}
create_long <- function(input_df, sam_tab = sam_table_mi,
                        tax_tab = tax_table_mi) {
  input_df %>%
    as_tibble() %>%
    mutate(species = tax_table_mi$species) %>%
    relocate(species) %>%
    pivot_longer(cols = starts_with("X"),
                 names_to = "Sample", values_to = "Abundance") %>%
    left_join(., sam_tab, by = "Sample") %>%
    left_join(., tax_tab, by = "species") %>%
    relocate(colnames(sam_tab))
}

alluvial_df <- create_long(counts_table_mi)

``` 

### Abundances of different taxons

```{R percent abundance}

# Non-limited to any groupings, all samples!
alluvial_df %>%
  group_by(genus) %>%
  summarise(pct = round(sum(Abundance)/sum(alluvial_df$Abundance)*100, 4)) %>%
  arrange(desc(pct))

alluvial_df %>%
  group_by(phylum) %>%
  summarise(pct = round(sum(Abundance)/sum(alluvial_df$Abundance)*100, 4)) %>%
  arrange(desc(pct))
```



## Beta Diversity Testing

### Genus - Infant
```{R}
# Genus
diversity_beta_test(subsetByColData(OG_dat,
                               which(OG_dat$MothChild == "Infant" &
                                     OG_dat$timepoint == 0)),
                    tax_level = "genus",
                    input_beta_method = "bray", #jaccard
                    input_select_beta_condition = "HIVStatus",
                    input_select_beta_stat_method = "Wilcoxon rank sum test")
```

```{R}
diversity_beta_test(subsetByColData(OG_dat,
                               which(OG_dat$MothChild == "Infant" &
                                     OG_dat$timepoint == 6)),
                    tax_level = "genus",
                    input_beta_method = "bray", #jaccard
                    input_select_beta_condition = "HIVStatus",
                    input_select_beta_stat_method = "Wilcoxon rank sum test")

#animalcules::diversity_beta_boxplot(subsetByColData(OG_dat,
#                               which(OG_dat$MothChild == "Infant")),
#                    tax_level = "genus",
#                    input_beta_method = "bray", #jaccard
#                    input_select_beta_condition = "HIVStatus")
```

### Genus - Mother
```{R}
# Genus
diversity_beta_test(subsetByColData(OG_dat,
                               which(OG_dat$MothChild == "Mother" &
                                     OG_dat$timepoint == 0)),
                    tax_level = "genus",
                    input_beta_method = "bray", #jaccard 
                    input_select_beta_condition = "HIVStatus",
                    input_select_beta_stat_method = "Wilcoxon rank sum test")
```

```{R}
diversity_beta_test(subsetByColData(OG_dat,
                               which(OG_dat$MothChild == "Mother" &
                                     OG_dat$timepoint == 6)),
                    tax_level = "genus",
                    input_beta_method = "bray", #jaccard 
                    input_select_beta_condition = "HIVStatus",
                    input_select_beta_stat_method = "Wilcoxon rank sum test")


#animalcules::diversity_beta_boxplot(subsetByColData(OG_dat,
#                               which(OG_dat$MothChild == "Mother")),
#                    tax_level = "genus",
#                    input_beta_method = "bray", #jaccard
#                    input_select_beta_condition = "HIVStatus")
```

### Genus - HEU infants t=0 vs t=6
```{R}
# Genus
diversity_beta_test(subsetByColData(OG_dat,
                               which(OG_dat$MothChild == "Infant" &
                                     OG_dat$HIVStatus == "HIV" &
                                     OG_dat$timepoint %in% c(0, 6))),
                    tax_level = "genus",
                    input_beta_method = "bray", #jaccard 
                    input_select_beta_condition = "timepoint",
                    input_select_beta_stat_method = "Wilcoxon rank sum test")

#animalcules::diversity_beta_boxplot(subsetByColData(OG_dat,
#                               which(OG_dat$MothChild == "Mother")),
#                    tax_level = "genus",
#                    input_beta_method = "bray", #jaccard
#                    input_select_beta_condition = "HIVStatus")
```


### Genus - HUU infants t=0 vs t=6
```{r}
diversity_beta_test(subsetByColData(OG_dat,
                               which(OG_dat$MothChild == "Infant" &
                                     OG_dat$HIVStatus == "Control" &
                                     OG_dat$timepoint %in% c(0, 6))),
                    tax_level = "genus",
                    input_beta_method = "bray", #jaccard 
                    input_select_beta_condition = "timepoint",
                    input_select_beta_stat_method = "Wilcoxon rank sum test")

```

## Plotting

### Obtain most abundant genera
```{R}
obtain_gn <- function(counts_tab, tax_tab) {
  all_relabu_genus <- counts_tab |>
    as.matrix() |>
    # Get rel abu within samokes
    prop.table(margin = 2) |>
    as_tibble() |>
    bind_cols(genus = tax_tab$genus) |>
    relocate(genus) |>
    group_by(genus) |>
    # Sum rel abu within samples/columns for genera
    summarise(across(.fns = sum)) %>%
    # Sum everything but the first columm ("genus")
    mutate(allmeans = apply(.[,-1], 1, mean)) |>
    select(genus, allmeans) |>
    mutate(genus = replace(genus, is.na(genus), "Unknown")) |>
    arrange(desc(allmeans))
  
  all_relabu_genus |>
    select(genus) |> unlist() |> unname() %>% .[1:20] %>%
    return()
}

ind <- microbe_mi$MothChild == "Infant"
best_genus <- obtain_gn(counts_table_mi[, ind], tax_table)
ind <- microbe_mi$MothChild == "Mother"
best_genus_m <- obtain_gn(counts_table_mi[, ind], tax_table_mi)
```

### Infants
```{r barplots control genus}
# Barplot for Control
p <- relabu_barplot(subsetByColData(OG_dat,
                               which(OG_dat$HIVStatus == "Control" &
                                       OG_dat$MothChild == "Infant")),
               tax_level = "genus",
               order_organisms = best_genus[1:20],
               sort_by = "conditions",
               group_samples = TRUE,
               group_conditions = "timepoint",
               show_legend = TRUE)

#plotly::export(p, file = "PaperFigs/RAW_barplot_control_inf.png",
#                   vwidth = 750, vheight = 550)
p
```

```{r barplots HIV-E genus}
# Barplot for Control
p <- relabu_barplot(subsetByColData(OG_dat,
                               which(OG_dat$HIVStatus == "HIV" &
                                       OG_dat$MothChild == "Infant")),
               tax_level = "genus",
               order_organisms = best_genus[1:20],
               sort_by = "conditions",
               group_samples = TRUE,
               group_conditions = "timepoint",
               show_legend = TRUE)

#plotly::export(p, file = "PaperFigs/RAW_barplot_HIV_inf.png",
#                   vwidth = 750, vheight = 550)
p
```

### Mothers
```{r barplots control genus moms}
# Barplot for Control
p <- relabu_barplot(subsetByColData(OG_dat,
                               which(OG_dat$HIVStatus == "Control" &
                                       OG_dat$MothChild == "Mother")),
               tax_level = "genus",
               order_organisms = best_genus_m[1:20],
               sort_by = "conditions",
               group_samples = TRUE,
               group_conditions = "timepoint",
               show_legend = TRUE)

#plotly::export(p, file = "PaperFigs/RAW_barplot_control_mom.png",
#                   vwidth = 750, vheight = 550)
p
```

```{r barplots HIV-E genus mom}
# Barplot for Control
p <- relabu_barplot(subsetByColData(OG_dat,
                               which(OG_dat$HIVStatus == "HIV" &
                                       OG_dat$MothChild == "Mother")),
               tax_level = "genus",
               order_organisms = best_genus_m[1:20],
               sort_by = "conditions",
               group_samples = TRUE,
               group_conditions = "timepoint",
               show_legend = TRUE)

#plotly::export(p, file = "PaperFigs/RAW_barplot_HIV_mom.png",
#                   vwidth = 750, vheight = 550)
p
```



