---
title: "Microbiome analysis of HEU and HUU Zambian infants"
author: "Aubrey R. Odom-Mabey"
date: "2022"
output:
  html_document:
    toc: true
    theme: united
    toc_depth: 6
    number_sections: true
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---
  
# Setup
```{r setup, message = FALSE}
suppressPackageStartupMessages({
  # General use
  library(tidyverse)
  library(knitr)
  library(magrittr)
  # Data formats
  library(animalcules)
  library(MultiAssayExperiment)
  library(SummarizedExperiment)
  # Modeling
  library(lme4)
  library(geepack)
  # Plotting
  library(alluvial)
  library(ggeffects)
  library(emmeans)
})
```

# Read in data
```{r plotting}
# Read in SE as "dat"
dat <- readRDS("data/FinalDatOther.rds")

# Infants only ----------------------------------------------------------------
microbe <- dat[, which(dat$MothChild == "Infant")]
sam_table <- as.data.frame(colData(microbe))

# Both mothers and infants ----------------------------------------------------
microbe_mi <- dat

# Extract metadata, taxonomic information, and counts
tax_table_mi <- as.data.frame(SummarizedExperiment::rowData(microbe_mi))
sam_table_mi <- as.data.frame(SummarizedExperiment::colData(microbe_mi))

counts_logcpm_mi <- SummarizedExperiment::assay(
  microbe_mi, "log_cpm") %>% as.data.frame()
counts_table_raw_mi <- SummarizedExperiment::assay(
  microbe_mi, "counts") %>% as.data.frame()

# Get raw counts tables
ind <- sam_table_mi$MothChild == "Infant"
counts_table_raw <- SummarizedExperiment::assay(microbe_mi, "counts") %>%
    as.data.frame() %>%
  select(sam_table_mi$Sample[ind])

ind <- sam_table_mi$MothChild == "Mother"
counts_table_raw_m <- SummarizedExperiment::assay(microbe_mi, "counts") %>%
    as.data.frame() %>%
  select(sam_table_mi$Sample[ind])

```

# Basic Data Exploration

## Abundances of different genera
### Genus
```{R}
all_genera_sums <- animalcules::upsample_counts(
  counts_table_raw_mi, tax_table_mi, "genus") %>%
  rowSums()

pct <- all_genera_sums/sum(all_genera_sums)*100 %>%
  as.vector() 

as.data.frame(pct) %>% tibble::rownames_to_column("col") %>%
  arrange(desc(pct))
```

### Phylum
```{R}
ind <- tax_table_mi$species == "Other"
tax_table_mi$phylum[ind] <- "Other"

all_phyla_sums <- animalcules::upsample_counts(
  counts_table_raw_mi, tax_table_mi, "phylum") %>%
  rowSums()

pct <- all_phyla_sums/sum(all_phyla_sums)*100 %>%
  as.vector() 

as.data.frame(pct) %>% tibble::rownames_to_column("col") %>%
  arrange(desc(pct))
```

## How many unique phyla, genera, species (post-“othering”)
```{R}
sp_count <- tax_table_mi %>%
  # Remove "others"
  filter(species != "Other",
         !stringr::str_detect(species, "_sp.")) %>%
  distinct(species) %>%
  dplyr::summarise(`Number of species` = n())

# Number of genera 
ge_count <- tax_table_mi %>%
  distinct(genus) %>%
  filter(genus != "Others") %>%
  dplyr::summarise(`Number of genera` = n())

# Number of phyla prior to subsetting
ph_count <- tax_table_mi %>%
  distinct(phylum) %>%
  dplyr::summarise(`Number of phyla` = n())

# Final counts
bind_cols(ph_count, ge_count, sp_count)
```

## Obtain most abundant genera

Using average relative abundance
```{R}
obtain_gn <- function(counts_tab, tax_tab) {
  all_relabu_genus <- counts_tab |>
    upsample_counts(tax_table = tax_tab, higher_level = "genus") |>
    counts_to_relabu() |>
    rownames_to_column(var = "genus") |>
    filter(genus != "Other") |>
    rowwise(genus) |>
    # Sum evergything but the first columm ("genus")
    summarise(allmeans = mean(c_across(starts_with("X"))),
              .groups = "drop") |>
    arrange(desc(allmeans))
  
  all_relabu_genus |> select(genus) |> unlist() |> unname() %>% 
    c(., "Other") %>% return()
}

ind <- microbe_mi$MothChild == "Infant"
best_genus <- obtain_gn(counts_table_raw_mi[, ind], tax_table_mi)

ind <- microbe_mi$MothChild == "Mother"
best_genus_m <- obtain_gn(counts_table_raw_mi[, ind], tax_table_mi)

# Both moms and infants
best_genus_mi <- obtain_gn(counts_table_raw_mi, tax_table_mi)
```

# Visualization & plots

## Stacked bar plots
```{r}
get_stacked_data <- function(SE, sam_tab, tax_tab, MI) {
  ind <- sam_tab$MothChild == MI
  sam_subset <- sam_tab[ind, ]
  counts_table_raw <- SummarizedExperiment::assay(SE, "counts") %>%
    as.data.frame() %>%
    select(sam_subset$Sample)
  relabu_table <- counts_table_raw %>%
    animalcules::upsample_counts(., tax_tab, "genus") %>%
    animalcules::counts_to_relabu() %>% t() %>% as.data.frame() %>%
    # Add grouping vars
    mutate(covariate1 = sam_subset$HIVStatus,
           covariate2 = sam_subset$timepoint) %>%
    pivot_longer(!c(covariate1, covariate2), names_to = "taxon") %>%
    S4Vectors::aggregate(. ~ taxon + covariate1 + covariate2, ., mean)
  return(relabu_table)
}

stacked_mom <- get_stacked_data(microbe_mi, sam_tab = sam_table_mi,
                                tax_tab = tax_table_mi, MI = "Mother")
stacked_inf <- get_stacked_data(microbe_mi, sam_tab = sam_table_mi,
                                tax_tab = tax_table_mi, MI = "Infant")
```

### Plot infants
```{R}
# Color palette
usepalette <- paletteer::palettes_d[["ggsci"]]["category20_d3"] |>
  unlist() |> unname() %>% 
  # Make "Other" show as grey
  .[seq_along(best_genus_mi[-1])] %>% c(., "grey")

# Plot stacked barplot
myplot <- stacked_inf %>%
  mutate("Genus" = factor(taxon, levels = best_genus_mi),
         "Age (days)" = factor(covariate2,
                               levels = seq(0, 6),
                               labels = c("0-15", "16-31", "32-47",
                                          "48-63", "64-79", "80-95", "96-120")),
         "HIV status" = factor(covariate1, levels = c("HIV", "Control"),
                               labels = c("HEU", "HUU"))) %>%
  dplyr::rename("Relative abundance" = `value`) %>%
  ggplot(aes(fill = Genus, x = `Age (days)`, y = `Relative abundance`)) + 
    geom_bar(position = "stack", stat = "identity") +
  facet_grid(~`HIV status`) + theme_classic() +
  labs(title = "Relative abundances of genera",
       subtitle = "HEU vs. HUU infants") +
  theme(legend.position = "bottom",
        #axis.title.x = element_blank(), axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  scale_fill_manual(values = usepalette) +
  coord_flip()
print(myplot)

ggsave(filename = "PaperFigs/stacked_barplot_infants.png",
  device = "png", width = 10, height = 6, units = "in", scale = 1)
```

### Plot mothers
```{R}
# Use ordered best_genus to match colors between plots

# Color palette
usepalette <- paletteer::palettes_d[["ggsci"]]["category20_d3"] |>
  unlist() |> unname() %>% 
  # Make "Other" show as grey
  .[seq_along(best_genus_mi[-1])] %>% c(., "grey")

# Plot stacked barplot
myplot <- stacked_mom %>%
  mutate("Genus" = factor(taxon, levels = best_genus_mi),
         "Time point" = factor(covariate2,
                               levels = c(0, 6),
                               labels = c("First", "Last")),
         "HIV status" = factor(covariate1, levels = c("HIV", "Control"),
                               labels = c("HIV+", "HIV-"))) %>%
  dplyr::rename("Relative abundance" = `value`) %>%
  ggplot(aes(fill = Genus, x = `Time point`, y = `Relative abundance`)) + 
    geom_bar(position = "stack", stat = "identity") +
  facet_grid(~`HIV status`) + theme_classic() +
  labs(title = "Relative abundances of genera",
       subtitle = "HIV+ vs. HIV- mothers") +
  theme(legend.position = "bottom",
        #axis.title.x = element_blank(), axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  scale_fill_manual(values = usepalette) +
  coord_flip()
print(myplot)

ggsave(filename = "PaperFigs/stacked_barplot_mothers.png",
  device = "png", width = 10, height = 6, units = "in", scale = 1)
```

### Finding a good palette
```{R, eval = FALSE}
#color_change <- function(num) {
#  myplot <- relabu_table %>%
#    mutate("Genus" = factor(taxon, levels = c("Other", taxon_order)),
#           "Age (days)" = factor(covariate2,
#                                 levels = seq(0, 6),
#                                 labels = c("0-15", "16-31", "32-47",
#                                            "48-63", "64-79", "80-95", "96-120")),
#           "HIV status" = factor(covariate1, levels = c("HIV", "Control"),
#                                 labels = c("HEU", "HUU"))) %>%
#    dplyr::rename("Relative abundance" = `value`) %>%
#    ggplot(aes(fill = Genus, x = `Age (days)`, y = `Relative abundance`)) + 
#      geom_bar(position = "stack", stat = "identity") +
#    facet_grid(~covariate1) +
#    theme_classic() +
#    labs(title = "Relative abundances of genera",
#         subtitle = names(colorcombo[num])) +
#    theme(legend.position = "bottom",
#          #axis.title.x = element_blank(), axis.text.x = element_blank(),
#          axis.ticks.x = element_blank()) +
#    scale_fill_manual(values = unname(unlist(colorcombo[num])))
#    coord_flip()
#  print(myplot)
#}
#
##ggsave(filename = paste0(
##  stem, "aubrey/Tyler_16S_paper/Plots/StackedKozich.png"),
##  device = "png", width = 12, height = 6, units = "in", scale = 1)
#
#all_palettes <- paletteer::palettes_d_names |>
#  filter(length > length(unique(relabu_table$taxon)),
#         type == "qualitative",
#         package == "ggsci") |>
#  select(package, palette)
#
#get_cols <- function(num) {
#  this_package <- all_palettes$package[num]
#  this_palette <- all_palettes$palette[num]
#  out <- paletteer::palettes_d[[this_package]][this_palette] |>
#    unlist() |> unname()
#  return(c("grey", out))
#}
#
#colorcombo <- lapply(seq_len(nrow(all_palettes)), get_cols)
#names(colorcombo) <- paste(all_palettes$package, all_palettes$palette)
#sapply(seq_along(colorcombo), color_change)
```

## Stacked area chart

```{r function for stacked barplot}
# Creates relative abundance table by timepoint, HIV status
get_relabu <- function(stacked_df) {
  relabu_table <- stacked_df %>%
  dplyr::rename(HIVStatus = "covariate1", timepoint = "covariate2", `Relative abundance` = "value",
         genus = taxon) %>%
  relocate(timepoint, HIVStatus, genus, `Relative abundance`)
  return(relabu_table)
}
```

```{R}
# Function to make plot
genus_stacked <- function(input_df, genus_inp, mothinf = "infant") {
  p <- input_df %>%
    mutate(Genus = factor(genus, levels = all_of(genus_inp))) %>%
    ggplot(aes(x = timepoint, y = `Relative abundance`,
               fill = Genus)) + 
    geom_area(alpha = 0.7, size = .5, colour = "white") +
    scale_fill_manual(values = usepalette) +
    theme_classic() +
    facet_grid(rows = vars(HIVStatus)) +
    labs(x = "Time point", y = "Relative abundance",
         title = paste("Microbe composition of", mothinf, "samples"),
         subtitle = "Relative abundance across samples")
  if(mothinf == "mother") return(p + scale_x_continuous(breaks = c(0, 6)))
  if(mothinf == "infant") return(p + scale_x_continuous(breaks = seq(0, 6)))
}
```

### Stacked area chart for infants
```{R}
input_df_inf <- get_relabu(stacked_inf)
genus_stacked(input_df_inf, best_genus_mi)
ggsave(filename = "PaperFigs/stacked_area_infants.png",
       width = 6, height = 4, units = "in",
       scale = 1)
```

### Stacked area chart for moms
```{R}
# Create plot
input_df_mom <- get_relabu(stacked_mom)
genus_stacked(input_df_mom, best_genus_mi, mothinf = "mother")
ggsave(filename = "PaperFigs/stacked_area_mothers.png",
       width = 6, height = 4, units = "in",
       scale = 1)
```

## Alluvial plot

```{R}
plot_alluvial <- function(input_df, MI, genus_inp, HIVpos, HIVneg) {
  p <- input_df %>%
    mutate(Genus = factor(genus, levels = genus_inp),
           HIVStatus = factor(HIVStatus, levels = c("HIV", "Control"), labels = c(HIVpos, HIVneg))) %>%
    # Get into alluvial format
    select(HIVStatus, timepoint, Genus, `Relative abundance`) %>%
    ggplot(aes(y = `Relative abundance`, x = timepoint, alluvium = Genus)) +
    ggalluvial::geom_alluvium(aes(fill = Genus, color = Genus),
                              width = 1/4, alpha = 0.7, decreasing = FALSE) +
    scale_fill_manual(values = usepalette) +
    scale_color_manual(values = usepalette) +
    theme_classic() +
    facet_grid(rows = vars(HIVStatus)) +
    labs(x = "Time point", y = "Relative abundance",
         title = "Relative abundance of genera over time",
         subtitle = MI)
  if(MI == "Mother") return(p + scale_x_continuous(breaks = c(0, 6)))
  if(MI == "Infant") return(p + scale_x_continuous(breaks = seq(0, 6)))
}
```

### Alluvial plot for infants
```{R}

plot_alluvial(input_df_inf, "Infant", genus_inp = best_genus_mi,
              HIVpos = "HEU", HIVneg = "HUU")
ggsave(filename = "PaperFigs/alluvial_genus_infant.png",
       width = 6, height = 4, units = "in",
       scale = 1)
```

### Alluvial plot for moms
```{R}
plot_alluvial(input_df_mom, "Mother", genus_inp = best_genus_mi,
              HIVpos = "HIV+", HIVneg = "HIV-")
ggsave(filename = "PaperFigs/alluvial_genus_mother.png",
       width = 6, height = 4, units = "in",
       scale = 1)
```

# Modeling

## Log CPM genus
```{R}
# Organize the data
logCPM_genus <- SummarizedExperiment::assay(microbe_mi, "counts") %>%
    as.data.frame() %>%
  animalcules::upsample_counts(tax_table_mi, "genus") %>%
  animalcules::counts_to_logcpm() %>%
  as.data.frame()
```

## Create long format dataframe

```{R create long}
create_long <- function(input_df, sam_tab = sam_table_mi) {
  input_df %>%
    rownames_to_column(var = "taxon") %>%
    pivot_longer(cols = starts_with("X"),
                 names_to = "Sample", values_to = "Abundance") %>%
    left_join(., sam_tab, by = "Sample") %>%
    return()
}
```

## Checking assumptions - histograms 

(Both infants/mothers)
```{R}
# Log CPM histogram
logCPM_genus %>%
  rownames_to_column(var = "genus") %>%
  pivot_longer(logCPM_genus, cols = starts_with("X"), names_to = "Sample") %>%
    ggplot(aes(x = value)) +
    geom_histogram(bins = 30) +
    labs(x = "Abundance", y = "log CPM",
         title = "Histogram of Abundance by Genus (Infants)",
         subtitle = "Log counts per million (CPM)") +
    facet_wrap(~genus) + 
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave(filename = "PaperFigs/hist_justification_logcpm.png",
       width = 5, height = 4, units = "in",
       scale = 1)

# Log CPM histogram
animalcules::upsample_counts(counts_table_raw, tax_table_mi, "genus")  %>%
  rownames_to_column(var = "genus") %>%
  pivot_longer(logCPM_genus, cols = starts_with("X"), names_to = "Sample") %>%
    ggplot(aes(x = value/10000)) +
    geom_histogram(bins = 30) +
    labs(x = "Abundance (tens of thousands)", y = "Count",
         title = "Histogram of abundance by genus (infants)",
         subtitle = "Counts/10000") +
    facet_wrap(~genus) + 
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave(filename = "PaperFigs/hist_justification_rawcounts.png",
       width = 5, height = 4, units = "in",
       scale = 1)
```

## Generalized Estimating Equations
- Run an independent GEE model for each taxa with relative abundance
- Works well with small data - multiple subpoints/subjects across clusters

https://data.library.virginia.edu/getting-started-with-generalized-estimating-equations/

### Functions to implement GEEs
```{R gee function}
test_models_gee <- function(tn, input_df, plot = FALSE) {
  filt_df <- input_df %>% filter(taxon == tn) %>%
    mutate(Subject = as.factor(Subject)) %>%
    arrange(Subject)
  complex <- geeglm(Abundance ~ 1 + HIVStatus + timepoint + HIVStatus*timepoint,
                    id = Subject, data = filt_df,
                    na.action = na.omit, family = "gaussian",
                    corstr = "ar1")
  sum_comp <- summary(complex)
  all_ci <- broom:::confint.geeglm(complex)
  if (plot) {
    # HIV Status, but timepoint is also visible
     p <- plot(ggeffects::ggemmeans(complex, terms = c("timepoint", "HIVStatus"))) + 
       ggplot2::labs(subtitle = "Estimated marginal means",
                     title = str_replace(tn, "_", " ")) +
       ylab("Taxon abundance (log CPM)") +
       xlab("Time point") +
      ggplot2::theme_classic()
     ggplot2::ggsave(filename = paste("PaperFigs/gee/", tn, ".png", sep = ""),
                     p, width = 6, height = 4, units = "in", scale = 0.7)
    return()
  }
  return(c(timepoint_p = sum_comp$coefficients$`Pr(>|W|)`[3],
           time_beta = sum_comp$coefficients$Estimate[3],
           time_beta_2.5 = all_ci["timepoint", "lwr"],
           time_beta_97.5 = all_ci["timepoint", "upr"],
           HIVStatus_p = sum_comp$coefficients$`Pr(>|W|)`[2],
           HIV_beta = sum_comp$coefficients$Estimate[2],
           HIV_beta_2.5 = all_ci["HIVStatusHIV", "lwr"],
           HIV_beta_97.5 = all_ci["HIVStatusHIV", "upr"],
           interaction_p = sum_comp$coefficients$`Pr(>|W|)`[4],
           interaction_beta = sum_comp$coefficients$Estimate[4],
           interaction_beta_2.5 = all_ci["HIVStatusHIV:timepoint", "lwr"],
           interaction_beta_97.5 = all_ci["HIVStatusHIV:timepoint", "upr"]
           ))
}

save_table <- function(input_df = results_species,
                       fileout = "PaperFigs/species_us_gee.csv") {
  input_df %>%
  select("Timepoint Unadjusted p-values" = timepoint_p,
         "Timepoint Coefficient Estimate" = time_beta,
         "Timepoint Adjusted p-values" = adj_timepoint,
         "HIV Status Unadjusted p-values" = HIVStatus_p, 
         "HIV Status Coefficient Estimate" = HIV_beta,
         "HIV Status Adjusted p-values" = adj_HIVStatus,
         "Interaction Unadjusted p-values" = interaction_p, 
         "Interaction Coefficient Estimate" = interaction_beta,
         "Interaction Adjusted p-values" = adj_interaction) %>%
  round(., 4) %>%
  write.csv(., fileout)
}

plot_hist <- function(input_vec, ...) {
  qplot(input_vec, geom = "histogram", bins = 30) + 
    labs(y = "Count", title = "Histogram of p-values",
         x = "p-values", ...) +
    theme_bw()
}
```

```{R}
run_model <- function(logcpm_tab, to_plot = TRUE) {
    input_df <- logcpm_tab %>%
    filter(MothChild == "Infant") %>%
    group_by(Subject, HIVStatus, timepoint, taxon) %>% 
    dplyr::summarise(Abundance = sum(Abundance), .groups = "drop") %>%
    arrange(timepoint)
  all_tn <- unique(input_df$taxon)
  n <- length(all_tn)
  
  # Plot model coefficients
    # If emmeans doesn't work, may need to step through function
    # and make one plot
  if (to_plot) lapply(all_tn, test_models_gee, input_df = input_df, plot = TRUE)
  
  storage <- t(sapply(all_tn, test_models_gee, input_df = input_df)) %>%
    as_tibble %>%
    mutate(adj_timepoint = p.adjust(timepoint_p, method = "bonferroni"),
           adj_HIVStatus = p.adjust(HIVStatus_p, method = "bonferroni"),
           adj_interaction = p.adjust(interaction_p, method = "bonferroni")) %>%
    relocate(contains("HIV"), contains("time"), contains("interaction")) %>%
    as.data.frame() %>%
    magrittr::set_rownames(all_tn)
  
  #plot_hist(storage$HIVStatus, subtitle = "HIVStatus")
  #plot_hist(storage$timepoint, subtitle = "Timepoint")
  kable(storage, caption = 
          "P-values for the effect of time and HIV status on abundance",
        row.names = TRUE)
  return(storage)
}
```

### Modeling genera
```{R}
# Input: create long version of CPM (both moms and infants)
long_lcpm_genus <- create_long(logCPM_genus)

# Get genus-level results
results_genus <- run_model(long_lcpm_genus)
save_table(results_genus, "PaperFigs/genus_ar1_gee.csv")

# Bonferroni cutoff - genus
0.05/nrow(results_genus)
```


### Modeling species
```{R}
# Input: create long version of CPM (both moms and infants)
  # This is filtered to infants only when modeling
long_lcpm_species <- create_long(counts_logcpm_mi)

# Top species (more than 1%)
filt <- animalcules::counts_to_relabu(counts_table_raw) |>
  apply(1, function(x) round(mean(x), 3)) %>%
  .[. > 0.01]
  ## Remove any species that are general "sp." or "Other"
ind <- stringr::str_detect(names(filt), "_sp.|Other")
species_to_test <- names(filt[!ind])

# Get species-level results
results_species <- long_lcpm_species %>%
  filter(taxon %in% species_to_test) %>%
  run_model()
save_table(results_species, "PaperFigs/species_ar1_gee.csv")

# Bonferroni cutoff - genus
0.05/nrow(results_species)
```

# Testing

## Number of samples available per timepoint
```{R paired infants Hotelling}
# Hotelling test for infant pairings (HIV vs. Controls) by species Abundance
# Note: some infants are missing timepoints

# Have to enter this manually
matched <- matrix(c("0554", "0469",
                    "0620", "0507",
                    "0645", "0539",
                    "0749", "0718",
                    "0798", "0801",
                    "0850", "0839",
                    "0895", "0878",
                    "0915", "0890",
                    "0987", "1040",
                    "1095", "1049"),
                  dimnames = list(c(), c("HIV", "Control")),
                  byrow = TRUE, ncol = 2)

# Add pairings to table
sam_table_mi_groups <- matched %>%
  as_tibble() %>%
  mutate(pairing = seq_len(nrow(matched))) %>%
  pivot_longer(cols = c(HIV, Control), names_to = c("HIVStatus"),
               values_to = "Subject") %>%
  mutate(Subject = paste(Subject, "-1", sep = "")) %>%
  full_join(sam_table_mi, ., by = c("Subject", "HIVStatus"))
```

### How many timepoints available per subject?
```{R}
sam_table_mi_groups %>% 
  filter(MothChild == "Infant") %>%
  group_by(timepoint, pairing) %>%
  dplyr::summarise(out = n()) %>%
  mutate(new = (out == 2)) %>% # num infant pairs
  dplyr::summarise(sum(new))
sam_table_mi_groups %>% 
  filter(MothChild == "Mother") %>%
  group_by(timepoint) %>%
  dplyr::summarise(n())
```

### Which infants/moms are missing timepoints?
```{R}
sam_table %>%
  group_by(Subject) %>%
  summarize(count = n()) %>%
  group_by(count) %>%
  summarize(n())

# A tibble: 4 × 2
# num timepoints || how many infants
# <int> <int>
# 2     1
# 3     1
# 6     2
# 7    16
```

### How many infants per timepoint?
```{R}
sam_table %>%
  group_by(timepoint) %>%
  summarize(count = n()) 
```

### How many mothers per timepoint?
```{R}
sam_table_mi %>%
  filter(MothChild == "Mother") %>%
  group_by(Subject) %>%
  summarize(count = n()) %>%
  group_by(count) %>%
  summarize(n())
```

## Paired multivariate Hotelling's T-squared

```{r Hotelling main function}
# Hotelling's T^2 Multivariate function
# Source: https://online.stat.psu.edu/stat505/lesson/7/7.1/7.1.4
  # Example: https://online.stat.psu.edu/stat505/lesson/7/7.1/7.1.9
  # Null hypothesis: population mean vectors are equal
  #  Alternative: mean vectors are not equal

Hotelling_mv_T2 <- function(tmp, Group1, Group2){
  # Rename groups
  tmp %<>% select(pairing,
                  Taxon = "taxon",
                  Group1 = starts_with(Group1),
                  Group2 = starts_with(Group2))
  # Define n
  all_pairs <- unique(tmp$pairing)
  n <- length(all_pairs)
  p <- length(unique(tmp$Taxon))
  # Sample mean vector
  Y_i <- tmp %>%
    group_by(pairing, Taxon) %>%
    dplyr::summarise(Yi = Group1 - Group2,
                     .groups = "drop")
  Ybar <- Y_i %>%
    group_by(Taxon) %>%
    dplyr::summarise(Ybar = mean(Yi),
                     .groups = "drop")
  # Y_i - Ybar
  diff <- Y_i %>%
    left_join(., Ybar, by = "Taxon") %>%
    group_by(pairing, Taxon) %>%
    dplyr::summarise(diff = Yi - Ybar, .groups = "drop")
  # (y_i-Ybar)%*%(y_i-Ybar)'
  mult_func <- function(x) {
    vec <- t(t(diff$diff[diff$pairing == x]))
    vec %*% t(vec)
  }
  # Calculate S_Y
  # Sample Var-cov matrix of vecs Y_i
  S_Y <- 1/(n - 1) * Reduce("+", lapply(all_pairs, mult_func))
  # Calculate T^2
  # T^2 = n*Ybar'*inv(S_Y)*Ybar
  Ybar_v <- t(t(Ybar$Ybar))
  T_2 <- n * t(Ybar_v) %*% solve(S_Y) %*% Ybar_v
  #F statistic
  # Dist according to F_p,n-p
  F_stat <- (n - p) / (p * (n - 1)) * T_2
  # Reject H0 at alpha if F-val exceeds
  df1 <- p
  df2 <- n - p
  crit_F <- qf(0.95, df1, df2)
  pval <- 1 - pf(c(F_stat), df1, df2)
  
  # Conduct a paired t-test
  # Following # 1 in R&C pg. 140
  if(pval < 0.05) {
    ttest_paired <- function(tax) {
      for_testing <- tmp %>% filter(Taxon == tax)
      out_test <- t.test(x = for_testing$Group1, y = for_testing$Group2,
                         alternative = "two.sided",
                         paired = TRUE, var.equal = TRUE)
      output <- list(t = as.numeric(out_test$statistic), df = as.numeric(out_test$parameter),
                  diff_means = as.numeric(out_test$estimate), CI_2.5 = as.numeric(out_test$conf.int[1]),
                  CI_97.5 = as.numeric(out_test$conf.int[2]), "p-value" = as.numeric(out_test$p.value))
      return(lapply(output, base::round, digits = 4))
    }
    results <- sapply(unique(tmp$Taxon), ttest_paired)
    results <- rbind(results, "adj p-value" = p.adjust(results["p-value", ],
                                                       method = "bonferroni"))
    write.csv(results, "PaperFigs/pairedttest_HIV_mothinf.csv")
    message("Results from t-tests:")
    print(results)
  }
  
  return(list(df1 = df1, df2 = df2, crit_F = crit_F,
              F_stat = F_stat[1], pvalue = pval))
}
```

### Paired mom-infant duads

Note that this is using genera...


```{R mothers vs infants}
conduct_hotelling_HIV_genus <- function(HIV_status, time = 0){
  message("Running tests for HIV status: ", HIV_status,
          " And timepoint: ", time)
  tmp <- create_long(logCPM_genus) %>%
    filter(timepoint == time,
           HIVStatus == HIV_status,
           # To avoid n < p, use top 5-6 species
           taxon %in% best_genus_mi[1:6]) %>%
    mutate(pairing = sapply(strsplit(Subject, "-"), function(x) x[1])) %>%
    select(pairing, MothChild, taxon, Abundance) %>%
    pivot_wider(., id_cols = c(pairing, taxon), values_from = Abundance,
                names_from = MothChild) %>%
    filter(!is.na(Mother) & !is.na(Infant)) %>%
    arrange(pairing)
  Hotelling_mv_T2(tmp, "Mother", "Infant")
}

```

#### t=0
```{R}
# time = 0
results_genus <- sapply(c("HIV", "Control"), conduct_hotelling_HIV_genus, time = 0)
colnames(results_genus) <- c("HIV", "Control")
results_genus
write.csv(results_genus, "PaperFigs/mv_hotelling_inf_moth.csv")
```

#### t=1
```{R}
# time = 6
results6 <- sapply(c("HIV", "Control"), conduct_hotelling_HIV_genus, time = 6)
colnames(results6) <- c("HIV", "Control")
results6
write.csv(results6, "PaperFigs/mv_hotelling_inf_moth6.csv")
```

### Paired infants at t=0,...,6

```{R}
# Function to run Hotelling's T^2 test
conduct_hotelling_Inf <- function(this_time){
  tmp <- create_long(logCPM_genus,
                     sam_tab = sam_table_mi_groups) %>%
    filter(timepoint == this_time,
           MothChild == "Infant",
           # To avoid n < p, use top 5-6 species
           taxon %in% best_genus[1:6]) %>%
    select(pairing, HIVStatus, taxon, Abundance) %>%
    pivot_wider(., id_cols = c(pairing, taxon),
                values_from = Abundance,
                names_from = HIVStatus) %>%
    filter(!is.na(HIV) & !is.na(Control)) %>%
    arrange(pairing)
  Hotelling_mv_T2(tmp, "HIV", "Control")
}
```

```{R}
# Obtain results for infants
results <- sapply(0:6, conduct_hotelling_Inf)
colnames(results) <- paste("Timepoint", 0:6)
numeric_results <- round(apply(results, 2, as.numeric), 4)
rownames(numeric_results) <- rownames(results)
write.csv(numeric_results, "PaperFigs/hotelling_infonly_timepoints.csv")

# Changing pvalues likely due to lack of time points (see critical value)

```


## Unpaired multivariate Hotelling's T-squared
```{R Hotelling unpaired}
Hotelling_mv_T2_up <- function(tmp, Populations, Subjects,
                            taxon = "taxon"){
  # Sources: https://online.stat.psu.edu/stat505/lesson/7/7.1/7.1.14
  # Also https://online.stat.psu.edu/stat505/lesson/7/7.1/7.1.15
  # Null hypothesis: population mean vectors are equal
  # Alternative: mean vectors are not equal
  
  # Rename groups
  tmp %<>% select(Populations = starts_with(Populations),
                  Subjects = starts_with(Subjects),
                  Taxon = starts_with(taxon),
                  Abundance)
  Group1 <- unique(tmp$Populations)[1]
  Sub1 <- tmp %>% filter(Populations == Group1) %>% 
    distinct(Subjects) %>% ungroup() %>% select(Subjects) %>% unlist()
  Group2 <- unique(tmp$Populations)[2]
  Sub2 <- tmp %>% filter(Populations == Group2) %>% 
    distinct(Subjects) %>% ungroup() %>% select(Subjects) %>% unlist()
  # Define n
  n <- tmp %>% group_by(Populations) %>% distinct(Subjects) %>% summarize(n())
  n1 <- n %>% filter(Populations == Group1) %>% select(`n()`) %>% as.numeric()
  n2 <- n %>% filter(Populations == Group2) %>% select(`n()`) %>% as.numeric()
  p <- length(unique(tmp$Taxon))
  # Sample mean vector
  X_i <- tmp %>%
    dplyr::rename(Xi = "Abundance")
  Xbar <- X_i %>%
    group_by(Populations, Taxon) %>%
    dplyr::summarise(Xbar = mean(Xi), .groups = "drop")
  # X_i - Xbar
  diff <- X_i %>%
    left_join(., Xbar, by = c("Populations", "Taxon")) %>%
    group_by(Subjects, Taxon) %>%
    dplyr::summarise(diff = Xi - Xbar, .groups = "drop")
  # (X_ij-Xbar_i)%*%(X_ij-Xbar_i)'
  mult_func <- function(x) {
    vec <- t(t(diff$diff[diff$Subjects == x]))
    vec %*% t(vec)
  }
  # Calculate S_p
  # Sample Var-cov matrix of vecs Y_i
  S_1 <- 1/(n1 - 1) * Reduce("+", lapply(Sub1, mult_func))
  S_2 <- 1/(n2 - 1) * Reduce("+", lapply(Sub2, mult_func))
  S_p <- ((n1 - 1) * S_1 + (n2 - 1) * S_2) / (n1 + n2 - 2)
  
  # T^2 = t(Xbar_1-Xbar_2) %*% {S_p(1/n1 + 1/n2)}^-1 %*% (Xbar_1-Xbar_2)
  Xbar_1 <- Xbar %>% filter(Populations == Group1) %>% select(Xbar)
  Xbar_2 <- Xbar %>% filter(Populations == Group2) %>% select(Xbar)
  meandiff <- t(t(Xbar_1 - Xbar_2))
  T_2 <- t(meandiff) %*% solve(S_p * (1 / n1 + 1 / n2)) %*% meandiff
  
  #F statistic
  # Dist according to F_p, n1 + n2 - p - 1
  F_stat <- (n1 + n2 - p - 1) / (p * (n1 + n2 - 2)) * T_2
  # Reject H0 at alpha if F-val exceeds
  df1 <- p
  df2 <- n1 + n2 - p - 1
  crit_F <- qf(0.95, df1, df2)
  pval <- 1 - pf(c(F_stat), df1, df2)
  
  if(pval < 0.05) {
    ttest_unpaired <- function(tax) {
      for_testing <- tmp %>% filter(Taxon == tax)
      out_test <- t.test(Abundance ~ Populations, for_testing,
                         alternative = "two.sided",
                         var.equal = FALSE, paired = FALSE)
      return(list(t = out_test$statistic, df = out_test$parameter,
                  diff_means = out_test$estimate[1],
                  CI_2.5 = out_test$conf.int[1],
                  CI_97.5 = out_test$conf.int[2],
                  "p-value" = out_test$p.value))
      }
    results <- sapply(unique(tmp$Taxon), ttest_unpaired)
    results <- rbind(results,
                     "adj p-value" = p.adjust(results["p-value", ],
                                              method = "bonferroni"))
    
    message("Results from t-tests:")
    print(results)
  }
  
  list(df1 = df1, df2 = df2, crit_F = crit_F,
       F_stat = F_stat[1], pvalue = pval)
}
```

### Unpaired moms comparison

```{R Hotelling mothers only}
conduct_hotelling_mother <- function(this_time, num_taxa){
  tmp <- create_long(logCPM_genus) %>%
    filter(MothChild == "Mother",
           # Avoid n > p
           taxon %in% best_genus_m[1:num_taxa],
           timepoint == this_time) %>%
    select(taxon, Subject, HIVStatus, Abundance) %>%
  arrange(Subject)
  return(Hotelling_mv_T2_up(tmp, "HIVStatus", "Subject", "taxon"))
}
```

#### t=0
```{R}
# Across all genera
results_moms0 <- conduct_hotelling_mother(0, 12)

results_moms0

write.csv(results_moms0, "PaperFigs/unpaired_moms_hotelling_t0.csv")
```

#### t=6
```{R}
# Across all genera
results_moms6 <- conduct_hotelling_mother(6, 12)

results_moms6

write.csv(results_moms6, "PaperFigs/unpaired_moms_hotelling_t6.csv")

# Biggest differences in staph, strep...
```

## Pathogens in paired moms-infants

```{R}
# Neiserria meningitidis is <0.1% (in "Other") so I won't test this

times <- c(0, 6)
hivstat <- c("HIV", "Control")
species_to_test <- c("Streptococcus_pneumoniae",
                     "Haemophilus_influenzae",
                     "Staphylococcus_haemolyticus")

init_table <- tibble(Species = rep(species_to_test,
                                   each = length(times) * length(hivstat)),
                     HIVstatus = rep(rep(hivstat,
                                         each = length(times)),
                       times = length(species_to_test)),
                     Timepoint = rep(times, length(species_to_test) *
                                       length(hivstat)))

testing_species <- function(which_species, which_status, which_time) {
  to_test <- counts_table_raw_mi %>%
  counts_to_logcpm() %>%  create_long() %>%
  filter(taxon == which_species,
         HIVStatus == which_status,
         timepoint == which_time) %>%
  select(Abundance, Subject, MothChild) %>%
  relocate(Abundance) %>%
  arrange(Subject) %>%
  tidyr::separate(Subject, sep = "-", into = c("Subject"), extra = "drop") %>%
  pivot_wider(., id_cols = Subject, id_expand = FALSE,
              names_from = MothChild, values_from = Abundance) 
  output <- t.test(to_test$Mother, to_test$Infant, paired = TRUE,
                   alternative = "two.sided")
  print(output)
  c("p value" = round(output$p.value, 2),
    "Lower CI" = round(output$conf.int[1], 2),
    "Upper CI" = round(output$conf.int[2], 2),
    "Mean of Differences in log counts" = round(unname(output$estimate), 2),
    "t" = round(unname(output$statistic), 2),
    "df" = round(unname(output$parameter), 2))
}
```


```{R}
# x - y!
# Conducts tests on mothers - infants

init_table %>%
  apply(., 1, function(x) testing_species(x[1], x[2], x[3])) %>%
  t() %>%
  as_tibble() %>%
  mutate("Adj P value" = p.adjust(`p value`, method = "bonferroni")) %>%
  bind_cols(init_table, .) %>%
  write.csv(., "PaperFigs/pathogens_pairedt.csv", row.names = FALSE)

```
